A pipeline to analyse time-course gene expression data

### 2019-07-22 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

\begin{itemize}
\tightlist
\item
  Add legends to the plots that don't have.
\item
  Currently have two functions ``plot\_genes'', ``plot\_centroids'', that do exactly
  the same thing. Find a better name for both of them
\item
  switch to camel case function names
\item
  Remove the namespace call
\item
  Make clear that it applies to RNASeq and how to call the different functions
  toRNASeq
\item
  Check that clustering can deal with count data.
\end{itemize}

@@

Gene expression studies provide simultaneous quantification of the level of
mRNA from all genes in a sample. High-throughput studies of gene expression
have a long history, starting with microarray technologies in the 1990s
through to single-cell technologies. While many expression studies are
designed to compare the gene expression in distinct groups, there is also a
long history of time-course expression studies, where the the gene expression
is compared across time by measuring mRNA levels from different samples across
time \footnote{Because the collection of the mRNA is often destructive, samples at
  different time points are generally from different biological samples;
  longitudinal studies, for example tracking the same subject over time, are
  certainly possible in certain settings, but not directly considered here.}. Such time course studies can vary from measuring a few distinct time
points, to sampling ten to twenty time points. Many longer time series are
particularly interested in investigating development over time. More recently,
single-cell studies track single cells through their development, and a single
cell is measured at a particular moment in its developmental progression -- a
value that is not know but estimated from the data as its ``pseudo-time.''

While there are many methods that have been proposed for discrete aspects of
time course data, the entire workflow for analysis of such data remains
difficult, particularly for long, developmental time series. Most methods
proposed for time course data are concerned with detecting genes that are
changing over time (differential expression analysis), examples being \texttt{edge}
\citep{storey:significance}, functional component analysis based models \citep{wu:more},
time-course permutation tests \citep{park:statistical}, and multiple testing
strategies to combine singe time point differential expression analysis
\citep{sun:multiple}. However, with long time course datasets, particularly in
developmental systems, a massive number of genes will show \emph{some} change. For
example, in the mice lung tissues infected with influenza, over 50\% of genes
are shown to be changing over time. The task in these settings is often not to
detect changes in genes, but to categorize into biologically interpretable
patterns the vast number of changes discovered.

We present here a workflow for such an analysis that consists of 4 main
parts (Figure \ref{fig:workflow}):

@@

@@

The first steps of analysis of gene expression data is always to do
normalization and quality control checks of the data. In what follows, we show
an example of this for the influenza data using generic methods; these steps
are not in specific to time course, but could be done for any gene expression
analysis.

First let's load the data. The package \texttt{moanin} contains the normalized
data and meta of \citep{shoemaker:ultrasensitive}.

@@

\CommentTok{# convert gene names}
\NormalTok{genes =}\StringTok{ }\KeywordTok{getBM}\NormalTok{(}\DataTypeTok{attributes=}\KeywordTok{c}\NormalTok{(}\StringTok{"ensembl_gene_id"}\NormalTok{, }\StringTok{"entrezgene"}\NormalTok{),}
        \DataTypeTok{filters=}\StringTok{"refseq_mrna"}\NormalTok{, }\DataTypeTok{values=}\NormalTok{genes,}
        \DataTypeTok{mart=}\NormalTok{ensembl)[}\StringTok{"entrezgene"}\NormalTok{]}
\NormalTok{genes =}\StringTok{ }\KeywordTok{as.vector}\NormalTok{(}\KeywordTok{unlist}\NormalTok{(genes))}
\NormalTok{pathways =}\StringTok{ }\KeywordTok{find_enriched_pathway}\NormalTok{(}
\NormalTok{    genes, }\DataTypeTok{species=}\StringTok{"mmu"}\NormalTok{,}
    \DataTypeTok{download_latest=}\OtherTok{TRUE}\NormalTok{)}\OperatorTok{$}\NormalTok{stastic}
\end{Highlighting}
\end{Shaded}

\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedleft}X>{\raggedleft}X}
\hline
  & Pathway & Percentage & Adj. p-value\\
\hline
05169 & Epstein-Barr virus infection & 37 & 0\\
\hline
04060 & Cytokine-cytokine receptor interaction & 31 & 0\\
\hline
05164 & Influenza A & 38 & 0\\
\hline
04668 & TNF signaling pathway & 45 & 0\\
\hline
04621 & NOD-like receptor signaling pathway & 34 & 0\\
\hline
05162 & Measles & 37 & 0\\
\hline
04630 & JAK-STAT signaling pathway & 33 & 0\\
\hline
05160 & Hepatitis C & 34 & 0\\
\hline
04620 & Toll-like receptor signaling pathway & 40 & 0\\
\hline
05167 & Kaposi sarcoma-associated herpesvirus infection & 29 & 0\\
\hline
\end{tabu}

@@

\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedleft}X>{\raggedleft}X>{\raggedleft}X>{\raggedright}X>{\raggedleft}X}
\hline
  & GO ID & Description & Annotated & Significant & Expected & P-value & adj. p-value\\
\hline
5 & GO:0003002 & regionalization & 172 & 153 & 134.46 & 0.00024 & 0.2416\\
\hline
6 & GO:0048562 & embryonic organ morphogenesis & 154 & 137 & 120.39 & 0.00035 & 0.2936\\
\hline
7 & GO:0010970 & transport along microtubule & 81 & 74 & 63.32 & 0.00057 & 0.4098\\
\hline
8 & GO:0060173 & limb development & 92 & 84 & 71.92 & 0.00070 & 0.4404\\
\hline
9 & GO:0060070 & canonical Wnt signaling pathway & 147 & 130 & 114.92 & 0.00090 & 0.5033\\
\hline
10 & GO:0055114 & oxidation-reduction process & 531 & 457 & 415.10 & 0.00135 & 0.6223\\
\hline
11 & GO:0019395 & fatty acid oxidation & 53 & 51 & 41.43 & 0.00136 & 0.6223\\
\hline
12 & GO:0006805 & xenobiotic metabolic process & 44 & 42 & 34.40 & 0.00168 & 0.7046\\
\hline
13 & GO:0090596 & sensory organ morphogenesis & 134 & 118 & 104.75 & 0.00224 & 0.8197\\
\hline
14 & GO:0010811 & positive regulation of cell-substrate ad... & 78 & 71 & 60.98 & 0.00228 & 0.8197\\
\hline
15 & GO:0060411 & cardiac septum morphogenesis & 33 & 32 & 25.80 & 0.00298 & 0.9999\\
\hline
16 & GO:0051056 & regulation of small GTPase mediated sign... & 151 & 131 & 118.04 & 0.00349 & 1.0000\\
\hline
17 & GO:0042738 & exogenous drug catabolic process & 22 & 22 & 17.20 & 0.00441 & 1.0000\\
\hline
18 & GO:1901381 & positive regulation of potassium ion tra... & 22 & 22 & 17.20 & 0.00441 & 1.0000\\
\hline
19 & GO:0007368 & determination of left/right symmetry & 67 & 61 & 52.38 & 0.00464 & 1.0000\\
\hline
\end{tabu}

@@

\begin{verbatim}
## R version 3.6.0 (2019-04-26)
## Platform: x86_64-pc-linux-gnu (64-bit)
## Running under: Ubuntu 19.04
## 
## Matrix products: default
## BLAS:   /usr/lib/x86_64-linux-gnu/blas/libblas.so.3.8.0
## LAPACK: /usr/lib/x86_64-linux-gnu/lapack/liblapack.so.3.8.0
## 
## locale:
## [1] C
## 
## attached base packages:
##  [1] stats4    parallel  splines   stats     graphics  grDevices utils    
##  [8] datasets  methods   base     
## 
## other attached packages:
##  [1] RColorBrewer_1.1-2   moanin_0.0.0         topGO_2.32.0        
##  [4] SparseM_1.77         GO.db_3.6.0          AnnotationDbi_1.42.1
##  [7] IRanges_2.14.12      S4Vectors_0.18.3     graph_1.58.2        
## [10] viridis_0.5.1        viridisLite_0.3.0    KEGGprofile_1.22.0  
## [13] RCurl_1.95-4.12      bitops_1.0-6         NMF_0.21.0          
## [16] Biobase_2.40.0       BiocGenerics_0.26.0  cluster_2.0.9       
## [19] rngtools_1.3.1.1     pkgmaker_0.27        registry_0.5-1      
## [22] ggplot2_3.1.1        biomaRt_2.36.1       BiocStyle_2.8.2     
## [25] kableExtra_1.1.0     knitr_1.22           limma_3.36.5        
## [28] usethis_1.5.0        devtools_2.0.2       rmarkdown_1.12      
## 
## loaded via a namespace (and not attached):
##  [1] colorspace_1.4-1        rprojroot_1.3-2        
##  [3] XVector_0.20.0          fs_1.3.1               
##  [5] rstudioapi_0.10         remotes_2.0.4          
##  [7] bit64_0.9-7             ClusterR_1.1.9         
##  [9] KEGG.db_3.2.3           xml2_1.2.0             
## [11] codetools_0.2-16        doParallel_1.0.14      
## [13] pkgload_1.0.2           ade4_1.7-13            
## [15] gridBase_0.4-7          png_0.1-7              
## [17] FD_1.0-12               readr_1.3.1            
## [19] compiler_3.6.0          httr_1.4.0             
## [21] backports_1.1.4         Matrix_1.2-17          
## [23] assertthat_0.2.1        lazyeval_0.2.2         
## [25] cli_1.1.0               htmltools_0.3.6        
## [27] prettyunits_1.0.2       tools_3.6.0            
## [29] gmp_0.5-13.5            gtable_0.3.0           
## [31] glue_1.3.1              reshape2_1.4.3         
## [33] dplyr_0.8.0.1           BiocWorkflowTools_1.6.2
## [35] Rcpp_1.0.1              Biostrings_2.48.0      
## [37] NMI_2.0                 ape_5.3                
## [39] nlme_3.1-139            iterators_1.0.10       
## [41] xfun_0.6                stringr_1.4.0          
## [43] ps_1.3.0                testthat_2.1.1         
## [45] rvest_0.3.3             gtools_3.8.1           
## [47] XML_3.98-1.19           zlibbioc_1.26.0        
## [49] MASS_7.3-51.4           scales_1.0.0           
## [51] hms_0.4.2               curl_3.3               
## [53] yaml_2.2.0              memoise_1.1.0          
## [55] gridExtra_2.3           TeachingDemos_2.10     
## [57] stringi_1.4.3           RSQLite_2.1.1          
## [59] desc_1.2.0              foreach_1.4.4          
## [61] permute_0.9-5           pkgbuild_1.0.3         
## [63] bibtex_0.4.2            geometry_0.4.1         
## [65] rlang_0.3.4             pkgconfig_2.0.2        
## [67] matrixStats_0.54.0      evaluate_0.13          
## [69] lattice_0.20-38         purrr_0.3.2            
## [71] bit_1.1-14              processx_3.3.1         
## [73] tidyselect_0.2.5        plyr_1.8.4             
## [75] magrittr_1.5            bookdown_0.9           
## [77] R6_2.4.0                DBI_1.0.0              
## [79] mgcv_1.8-28             pillar_1.3.1           
## [81] withr_2.1.2             abind_1.4-5            
## [83] KEGGREST_1.20.2         tibble_2.1.1           
## [85] crayon_1.3.4            progress_1.2.0         
## [87] grid_3.6.0              blob_1.1.1             
## [89] callr_3.2.0             git2r_0.25.2           
## [91] vegan_2.5-4             digest_0.6.18          
## [93] webshot_0.5.1           xtable_1.8-4           
## [95] munsell_0.5.0           magic_1.5-9            
## [97] sessioninfo_1.1.1
\end{verbatim}

>>>>

\begin{itemize}
\tightlist
\item
  Add legends to the plots that don't have.
\item
  Currently have two functions ``plot\_genes'', ``plot\_centroids'', that do exactly
  the same thing. Find a better name for both of them
\item
  switch to camel case function names
\item
  Remove the namespace call
\item
  Make clear that it applies to RNASeq and how to call the different functions
  to RNASeq
\item
  Check that clustering can deal with count data.
\end{itemize}

EAP:

\begin{itemize}
\tightlist
\item
  Do we need installation instructions for moanin in this document, since its not on CRAN/bioconductor? Also, how its set up now (doing make file, which compiles the documentation, etc), probably prohibits use of, e.g. \texttt{install\_github} from \texttt{devtools}. If able to call \texttt{install\_github("NelleV/2019timecourse-rnaseq-pipeline",\ subdir="bin/moanin",\ dependencies=TRUE)} would be simpler (would also install the dependencies and get around problems in script)
\item
  I got an error in installation script for packages:
\end{itemize}

\begin{verbatim}
epurdom-0:2019timecourse-rnaseq-pipeline epurdom$ Rscript scripts/install.R
Loading required package: devtools
Loading required package: usethis
Loading required package: ggfortify
Loading required package: ggplot2
Loading required package: kableExtra
Error in contrib.url(repos, "source") : 
  trying to use CRAN without setting a mirror
Calls: install.packages -> contrib.url
In addition: Warning message:
In library(package, lib.loc = lib.loc, character.only = TRUE, logical.return = TRUE,  :
  there is no package called <e2><80><98>kableExtra<e2><80><99>
Execution halted
\end{verbatim}

\begin{itemize}
\item
  \texttt{biocLite} is no longer the recommended way to install bioconductor. Instead \texttt{BiocManager::install(c("edge",\ "NMI"))}. But this is dependent on bioconductor/R version. Also, now that docker, can you change the installation to just be simple one-liner (and \texttt{BiocManager::install} can also do non-bioc packages, and is much more reliable.)
\item
  I get multiple errors in compilation of file. Not sure, however, why TravisCI didn't hit these problems, so you might want to see what's going on there. Probably caching or something, so might want to occassionally remove the cache and run again, if TravisCI is saving the cache. One error I fixed, but wasn't sure about this error (changed it to \texttt{entrezgene\_id} and seems to be running now\ldots{}):
\end{itemize}

\begin{verbatim}
label: enrichment_analysis (with options) 
List of 2
 $ results: chr "hide"
 $ message: logi FALSE

Quitting from lines 1124-1138 (manuscript.Rmd) 
Error in getBM(attributes = c("ensembl_gene_id", "entrezgene"), filters = "refseq_mrna",  : 
  Invalid attribute(s): entrezgene 
Please use the function 'listAttributes' to get valid attribute names
Calls: <Anonymous> ... withCallingHandlers -> withVisible -> eval -> eval -> getBM
\end{verbatim}

@@

Gene expression studies provide simultaneous quantification of the level of
mRNA from all genes in a sample. High-throughput studies of gene expression
have a long history, starting with microarray technologies in the 1990s
through to single-cell technologies. While many expression studies are
designed to compare the gene expression between distinct groups, there is also a
long history of time-course expression studies. Such studies compare the gene expression
across time by measuring mRNA levels from samples collected at different timepoints \footnote{Because the collection of the mRNA is often destructive, samples at
  different time points are generally from different biological samples;
  longitudinal studies, for example tracking the same subject over time, are
  certainly possible, but not directly considered here.}. Such time-course studies can vary from measuring a few distinct time
points, to sampling ten to twenty time points. These longer time series are
particularly interested in investigating development over time. More recently,
a new variety of time course studies have come from single-cell sequencing experiments\textbf{EAP: need citation} which can sequencing single cells which are at different stages of development; in this case, the time point is the stage of the cell in the process of development -- a
value that is not know but estimated from the data as its ``pseudo-time.''

While there are many methods that have been proposed for discrete aspects of
time course data, the entire workflow for analysis of such data remains
difficult, particularly for long, developmental time series. Most methods
proposed for time course data are concerned with detecting genes that are
changing over time (differential expression analysis), examples being \texttt{edge}
\citep{storey:significance}, functional component analysis based models \citep{wu:more},
time-course permutation tests \citep{park:statistical}, and multiple testing
strategies to combine single time point differential expression analysis
\citep{sun:multiple}. However, with long time course datasets, particularly in
developmental systems, a massive number of genes will show \emph{some} change. For
example, in a study of mice lung tissues infected with influenza that we consider in this workflow, over 50\% of genes
are shown to be changing over time. The task in these settings is often not to
detect changes in genes, but to categorize them into biologically interpretable
patterns.

We present here a workflow for such an analysis that consists of 4 main
parts (Figure \ref{fig:schema}):

@@

\textbf{EAP: We need to add comment about microarray versus sequencing data}

@@

The first steps of analysis of gene expression data is always to do
normalization and quality control checks of the data. In what follows, we show
an example of this for the influenza data using generic methods; these steps
are not in specific to time course data, but could be done for any gene expression
analysis.

First let's load the data. The package \texttt{moanin} contains the normalized
data and metadata of \citep{shoemaker:ultrasensitive}.

@@

\CommentTok{# convert gene names}
\NormalTok{genes =}\StringTok{ }\KeywordTok{getBM}\NormalTok{(}\DataTypeTok{attributes=}\KeywordTok{c}\NormalTok{(}\StringTok{"ensembl_gene_id"}\NormalTok{, }\StringTok{"entrezgene_id"}\NormalTok{),}
        \DataTypeTok{filters=}\StringTok{"refseq_mrna"}\NormalTok{, }\DataTypeTok{values=}\NormalTok{genes,}
        \DataTypeTok{mart=}\NormalTok{ensembl)[}\StringTok{"entrezgene_id"}\NormalTok{]}
\NormalTok{genes =}\StringTok{ }\KeywordTok{as.vector}\NormalTok{(}\KeywordTok{unlist}\NormalTok{(genes))}
\NormalTok{pathways =}\StringTok{ }\KeywordTok{find_enriched_pathway}\NormalTok{(}
\NormalTok{    genes, }\DataTypeTok{species=}\StringTok{"mmu"}\NormalTok{,}
    \DataTypeTok{download_latest=}\OtherTok{TRUE}\NormalTok{)}\OperatorTok{$}\NormalTok{stastic}
\end{Highlighting}
\end{Shaded}

\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedleft}X>{\raggedleft}X}
\hline
  & Pathway & Percentage & Adj. p-value\\
\hline
05169 & Epstein-Barr virus infection & 37 & 0\\
\hline
04060 & Cytokine-cytokine receptor interaction & 31 & 0\\
\hline
04668 & TNF signaling pathway & 45 & 0\\
\hline
05164 & Influenza A & 38 & 0\\
\hline
04621 & NOD-like receptor signaling pathway & 34 & 0\\
\hline
05162 & Measles & 37 & 0\\
\hline
04630 & JAK-STAT signaling pathway & 33 & 0\\
\hline
05160 & Hepatitis C & 34 & 0\\
\hline
04620 & Toll-like receptor signaling pathway & 40 & 0\\
\hline
05167 & Kaposi sarcoma-associated herpesvirus infection & 29 & 0\\
\hline
\end{tabu}

@@

\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedleft}X>{\raggedleft}X>{\raggedleft}X>{\raggedright}X>{\raggedleft}X}
\hline
  & GO ID & Description & Annotated & Significant & Expected & P-value & adj. p-value\\
\hline
5 & GO:0003002 & regionalization & 172 & 153 & 134.41 & 0.00026 & 0.2629\\
\hline
6 & GO:0048562 & embryonic organ morphogenesis & 152 & 135 & 118.78 & 0.00045 & 0.3323\\
\hline
7 & GO:0010970 & transport along microtubule & 82 & 75 & 64.08 & 0.00046 & 0.3323\\
\hline
8 & GO:0006805 & xenobiotic metabolic process & 49 & 47 & 38.29 & 0.00059 & 0.3729\\
\hline
9 & GO:0060173 & limb development & 92 & 84 & 71.89 & 0.00068 & 0.3820\\
\hline
10 & GO:0019395 & fatty acid oxidation & 53 & 51 & 41.42 & 0.00134 & 0.6194\\
\hline
11 & GO:0055114 & oxidation-reduction process & 536 & 461 & 418.84 & 0.00139 & 0.6194\\
\hline
12 & GO:0060070 & canonical Wnt signaling pathway & 149 & 131 & 116.43 & 0.00147 & 0.6194\\
\hline
13 & GO:0090596 & sensory organ morphogenesis & 134 & 118 & 104.71 & 0.00219 & 0.8517\\
\hline
14 & GO:0060411 & cardiac septum morphogenesis & 33 & 32 & 25.79 & 0.00295 & 1.0000\\
\hline
15 & GO:0010811 & positive regulation of cell-substrate ad... & 81 & 73 & 63.30 & 0.00384 & 1.0000\\
\hline
16 & GO:0042738 & exogenous drug catabolic process & 22 & 22 & 17.19 & 0.00437 & 1.0000\\
\hline
17 & GO:1901381 & positive regulation of potassium ion tra... & 22 & 22 & 17.19 & 0.00437 & 1.0000\\
\hline
18 & GO:0007368 & determination of left/right symmetry & 66 & 60 & 51.57 & 0.00536 & 1.0000\\
\hline
19 & GO:0060078 & regulation of postsynaptic membrane pote... & 66 & 60 & 51.57 & 0.00536 & 1.0000\\
\hline
\end{tabu}

@@

\begin{verbatim}
## R version 3.6.0 (2019-04-26)
## Platform: x86_64-pc-linux-gnu (64-bit)
## Running under: Ubuntu 19.04
## 
## Matrix products: default
## BLAS:   /usr/lib/x86_64-linux-gnu/blas/libblas.so.3.8.0
## LAPACK: /usr/lib/x86_64-linux-gnu/lapack/liblapack.so.3.8.0
## 
## locale:
## [1] C
## 
## attached base packages:
##  [1] stats4    parallel  splines   stats     graphics  grDevices utils    
##  [8] datasets  methods   base     
## 
## other attached packages:
##  [1] RColorBrewer_1.1-2   moanin_0.0.0         topGO_2.32.0        
##  [4] SparseM_1.77         GO.db_3.6.0          AnnotationDbi_1.42.1
##  [7] IRanges_2.14.12      S4Vectors_0.18.3     graph_1.58.2        
## [10] viridis_0.5.1        viridisLite_0.3.0    KEGGprofile_1.22.0  
## [13] RCurl_1.95-4.12      bitops_1.0-6         NMF_0.21.0          
## [16] Biobase_2.40.0       BiocGenerics_0.26.0  cluster_2.0.9       
## [19] rngtools_1.3.1.1     pkgmaker_0.27        registry_0.5-1      
## [22] ggplot2_3.1.1        biomaRt_2.36.1       BiocStyle_2.8.2     
## [25] kableExtra_1.1.0     knitr_1.22           limma_3.36.5        
## [28] usethis_1.5.0        devtools_2.0.2       rmarkdown_1.12      
## 
## loaded via a namespace (and not attached):
##  [1] colorspace_1.4-1        rprojroot_1.3-2        
##  [3] XVector_0.20.0          fs_1.3.1               
##  [5] rstudioapi_0.10         remotes_2.0.4          
##  [7] bit64_0.9-7             ClusterR_1.1.9         
##  [9] KEGG.db_3.2.3           xml2_1.2.0             
## [11] codetools_0.2-16        doParallel_1.0.14      
## [13] pkgload_1.0.2           ade4_1.7-13            
## [15] gridBase_0.4-7          png_0.1-7              
## [17] FD_1.0-12               readr_1.3.1            
## [19] compiler_3.6.0          httr_1.4.0             
## [21] backports_1.1.4         Matrix_1.2-17          
## [23] assertthat_0.2.1        lazyeval_0.2.2         
## [25] cli_1.1.0               htmltools_0.3.6        
## [27] prettyunits_1.0.2       tools_3.6.0            
## [29] gmp_0.5-13.5            gtable_0.3.0           
## [31] glue_1.3.1              reshape2_1.4.3         
## [33] dplyr_0.8.0.1           BiocWorkflowTools_1.6.2
## [35] Rcpp_1.0.1              Biostrings_2.48.0      
## [37] NMI_2.0                 ape_5.3                
## [39] nlme_3.1-139            iterators_1.0.10       
## [41] xfun_0.6                stringr_1.4.0          
## [43] ps_1.3.0                testthat_2.1.1         
## [45] rvest_0.3.3             gtools_3.8.1           
## [47] XML_3.98-1.19           zlibbioc_1.26.0        
## [49] MASS_7.3-51.4           scales_1.0.0           
## [51] hms_0.4.2               curl_3.3               
## [53] yaml_2.2.0              memoise_1.1.0          
## [55] gridExtra_2.3           TeachingDemos_2.10     
## [57] stringi_1.4.3           RSQLite_2.1.1          
## [59] desc_1.2.0              foreach_1.4.4          
## [61] permute_0.9-5           pkgbuild_1.0.3         
## [63] bibtex_0.4.2            geometry_0.4.1         
## [65] rlang_0.4.0             pkgconfig_2.0.2        
## [67] matrixStats_0.54.0      evaluate_0.13          
## [69] lattice_0.20-38         purrr_0.3.2            
## [71] bit_1.1-14              processx_3.3.1         
## [73] tidyselect_0.2.5        plyr_1.8.4             
## [75] magrittr_1.5            bookdown_0.9           
## [77] R6_2.4.0                DBI_1.0.0              
## [79] mgcv_1.8-28             pillar_1.3.1           
## [81] withr_2.1.2             abind_1.4-5            
## [83] KEGGREST_1.20.2         tibble_2.1.1           
## [85] crayon_1.3.4            progress_1.2.0         
## [87] grid_3.6.0              blob_1.1.1             
## [89] callr_3.2.0             git2r_0.25.2           
## [91] vegan_2.5-4             digest_0.6.18          
## [93] webshot_0.5.1           xtable_1.8-4           
## [95] munsell_0.5.0           magic_1.5-9            
## [97] sessioninfo_1.1.1
\end{verbatim}
